Predicting Success of Motivational Interviews with Recurrent Neural Network and Probabilistic Model for Patient-Provider Communication Sequences

### 2017-09-18 paper-amia-summit-2018/amia_summit_2018.tex
update abstract and introduction

\section*{Introduction}
Data in the form of temporally ordered sequences of discrete or continuous observations (e.g., symbolic sequences such as notes in patient EHR, diagnostic codes, protein or DNA sequences or continuous time series, such as ECG measurements) arise in various domains of health informatics. For classifying temporally ordered sequenced, sequence classification methods are utilized. Sequence classification has a broad range of real word applications from genomics and health informatics to finance and anomaly detection. In genomics research, sequence classification is widely used to classify protein and text sequence data \cite{yakhnenko2005discriminatively} and detect the function of new proteins \cite{deshpande2002evaluation}. In health informatics, ECG measurements are considered as multi-dimensional time series and are used to classify individuals as healthy or having a heart disease \cite{wei2006semi}. Sequence classification is also used for anomaly detection such as abnormal access to systems \cite{lane1999temporal} and malware \cite{drew2017polymorphic}.

In general, there are three categories of classification methods for temporally ordered sequences, e.g.: feature-based, distance metric-based and model-based method. Feature-based classification methods transform a sequence into a feature vector and apply a standard supervised learning algorithm, such as support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature}. Shapelet \cite{ye2009time} and pattern \cite{kudenko1998feature, lesh1999mining} based techniques as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of standard classifiers as well. Distance-based methods measure the similarity between sequences of behavior codes to determine the quality of classification. The most commonly used distance function is Euclidian distance \cite{keogh2003need} with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data. The order of observations in a sequence is challenging to model using feature-based methods because the traditional bag-of-words representation of features does not account for the order of observation. At the same time, distance-based methods are more appropriate for time series data. Model-based classification methods involve building a model based on the dataset. In other words, information was learned from the dataset, and use that as a "model" to classify sequences.

@@

One of the model-based classification methods first creates a probabilistic model of a behavioral code sequence, such as Markov Chain (MC) and Hidden Markov Model \cite{rabiner1989tutorial} (HMM). Markov models have limitations when applied to sequence classification because they cannot preserve long-term dependencies in sequences. Similarly, the hidden states of HMMs follow a first-order Markov chain and can select only one hidden state at each time step. Therefore, Markov chain and HMMs are only used for passing one or two preceding step(s) among sequence steps\cite{kundu1988recognition}. On the other hand, Recurrent Neural Networks (RNNs) have no such limitations in preserving long-term dependencies among sequence steps due to their unlimited memory. It was shown that RNNs are very effective in sequence modeling with temporal data. Excellent performance in a wide range of sequence modeling problems with temporal data has been observed using RNNs \cite{nion2013handwritten, lipton2015learning, choi2016doctor}. A variant of RNN called Long Short-Term Memory was proposed \cite{graves2013speech} which solved the vanishing gradient problem \cite{bengio1993problem}. LSTM showed excellent performance in several domains, including speech recognition \cite{graves2013speech}, handwriting recognition \cite{nion2013handwritten} and health informatics \cite{lipton2015learning, choi2016doctor}. In Lipton et. al.\cite{lipton2015learning}, LSTM was used as a multilabel classification model to recognize patterns in multivariate time series of clinical measurements such as body temperature, heart rate, blood pressure, etc. LSTM was also effectively used, recently, for designing a model, called Doctor AI \cite{choi2016doctor}, to predict the diagnosis and medication categories for a subsequent visit. A further simplification and improvement of the LSTM model, called the Gated Recurrent Unit\cite{chung2014empirical} (GRU), was later proposed. These variants show markedly improved performance among all RNN variants and as such, this work focuses on these variants of RNN specifically.

In this paper, we test the applicability of state of the art models for the classification of patient-provider communication sequences in the context of clinical interviews. Specifically, we focus on the transcripts of Motivational Interviews (MI) with obese adolescents and their caregivers. Childhood obesity is a serious public health concern in the United States and worldwide. Recent estimates indicate that approximately one-third (31.8\%) of US children age 2-19 years are overweight and 16.9\% are obese \cite{ogden2012prevalence}. Adolescents who are obese likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}. One approach to effective obesity intervention is Motivational Interviewing, an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change. The goal of MI is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change. These statements, referred to as ``change talk'' (or CT), consistently predict actual behavior change\cite{apodaca2009mechanisms} that can be sustained for as long as 34 months\cite{walker2011influence} after an interview. However, the ability of counselors to consistently elicit this type of patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews. Since manual examination and sequential pattern analysis of transcripts is a very time-consuming process, tailoring of MI interventions to particular populations can take years. Therefore, there is a need for informatics-based methods to facilitate the development of effective MI-based interventions, in general, and theoretically-grounded computational models to explore the mechanisms of MI's efficacy, in particular.

@@

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.50\textwidth]{figures/code_embed.eps}
    \caption{\textbf{T-SNE diagram of behavior code drawn after 4000 steps with perplexity 30}.}
    \label{fig:code_embedding}
\end{figure}

@@

\subsection*{\textit{Most likely communication sequences}}
Table~\ref{tab:common_patterns} provides examples of typical patient-provider communication sequences that frequently appear in successful and unsuccessful motivational interviews. It can be seen that most successful sequence starts with a summary of the discussion or structured statement. After that, if adolescents express positive change talk, the counselor immediately reflects on that to reinforce adolescent's intrinsic motivation about behavior change. On the other hand, asking open- or closed-ended questions with eliciting negative change talk can lead to negative change talk, even in the cases when adolescents were showing a positive tendency in their previous communication. This observation can be explained by adolescents quickly confused when reiterating negative information that undermines their motivation. Analyzing such cases will allow the counselors to determine the negative information that can be discussed during the interviews.

>>>>

\section*{Introduction}
Data in the form of temporally ordered sequences of discrete or continuous observations (e.g., symbolic sequences such as notes in patient EHR, diagnostic codes, protein or DNA sequences or continuous time series, such as ECG measurements) arise in various domains of health informatics. For classifying temporally ordered sequences, sequence classification methods are utilized. Sequence classification has a broad range of real word applications from genomics and health informatics to finance and anomaly detection \cite{yakhnenko2005discriminatively, wei2006semi, lane1999temporal}. In genomics research, sequence classification is widely used to classify protein and text sequence data and detect the function of new proteins. In health informatics, ECG measurements are considered as multi-dimensional time series and are used to classify individuals as healthy or having a heart disease. Sequence classification is also used for anomaly detection such as abnormal access to systems and malware.

%Sequence classification has a broad range of real word applications from genomics and health informatics to finance and anomaly detection. In genomics research, sequence classification is widely used to classify protein and text sequence data \cite{yakhnenko2005discriminatively} and detect the function of new proteins \cite{deshpande2002evaluation}. In health informatics, ECG measurements are considered as multi-dimensional time series and are used to classify individuals as healthy or having a heart disease \cite{wei2006semi}. Sequence classification is also used for anomaly detection such as abnormal access to systems \cite{lane1999temporal} and malware \cite{drew2017polymorphic}.

In general, there are three categories of classification methods for temporally ordered sequences, e.g.: feature-based, distance metric-based and model-based method. Feature-based classification methods transform a sequence into a feature vector and apply a standard supervised learning algorithm, such as support vector machine \cite{leslie2004fast} or decision tree \cite{chuzhanova1998feature}. 
%Shapelet \cite{ye2009time} and pattern \cite{kudenko1998feature, lesh1999mining} based techniques as well as hierarchical approaches \cite{nallam2016effective} have been proposed instead of standard classifiers as well. 
Distance-based methods measure the similarity between sequences of behavior codes to determine the quality of classification. The most commonly used distance function is euclidian distance with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data.  
%The most commonly used distance function is Euclidian distance \cite{keogh2003need} with Dynamic Time Wrapping \cite{keogh2000scaling} used for more flexible matching in time series data. 
The order of observations in a sequence is challenging to model using feature-based methods because the traditional bag-of-words representation of features does not account for the order of observation. At the same time, distance-based methods are more appropriate for time series data. Model-based classification methods involve building a model based on the dataset. In other words, information was learned from the dataset, and use that as a "model" to classify sequences.

@@

One of the model-based classification methods is probabilistic model, which creates a model of behavioral code sequences, such as Markov Chain (MC) and Hidden Markov Model \cite{rabiner1989tutorial} (HMM). Markov models have limitations when applied to sequence classification because they cannot preserve long-term dependencies in sequences. Similarly, the hidden states of HMMs follow a first-order Markov chain and can select only one hidden state at each time step. Therefore, Markov chain and HMMs are usually used for passing one or two preceding step(s) among sequence steps\cite{kundu1988recognition}. On the other hand, Recurrent Neural Networks (RNNs) have no such limitations in preserving long-term dependencies among sequence steps due to their unlimited memory. It was shown that RNNs are very effective in a wide range of sequence modeling problems with temporal data\cite{nion2013handwritten, lipton2015learning, choi2016doctor}. A variant of RNN called Long Short-Term Memory was proposed \cite{graves2013speech} which solved the vanishing gradient problem \cite{bengio1993problem} of traditional RNN. LSTM showed excellent performance in several domains, including speech recognition \cite{graves2013speech}, handwriting recognition \cite{nion2013handwritten} and health informatics \cite{lipton2015learning, choi2016doctor}. In Lipton et. al.\cite{lipton2015learning}, LSTM was used as a multilabel classification model to recognize patterns in multivariate time series of clinical measurements such as body temperature, heart rate, blood pressure, etc. LSTM was also effectively used for designing a model, called Doctor AI \cite{choi2016doctor}, for predicting the diagnosis and medication categories of a subsequent visit. A further simplification and improvement of the LSTM model, called the Gated Recurrent Unit\cite{chung2014empirical} (GRU), was later proposed. These variants show markedly improved performance among all RNN variants and as such, this work focuses on these variants of RNN specifically.

In this paper, we test the applicability of state-of-the-art models for the classification of patient-provider communication sequences in the context of clinical interviews. Specifically, we focus on the transcripts of Motivational Interviews (MI) with obese adolescents and their caregivers. Childhood obesity is a serious public health concern in the United States and worldwide. Recent estimates indicate that approximately one-third (31.8\%) of US children age 2-19 years are overweight and 16.9\% are obese \cite{ogden2012prevalence}. Adolescents who are obese likely continue to be obese in adulthood and have a greater risk of heart disease, type 2 diabetes, stroke, cancer, and osteoarthritis \cite{general2010surgeon}. One approach to effective obesity intervention is Motivational Interviewing, an evidence-based counseling technique to increase intrinsic motivation and self-efficacy for health-related behavior change. The goal of MI is to encourage patients to explore their own desires, ability, reasons, need for and commitment to the targeted behavior change. These statements, referred to as ``change talk'' or CHT, consistently predict actual behavior change\cite{apodaca2009mechanisms} that can be sustained for as long as 34 months\cite{walker2011influence} after an interview. However, the ability of counselors to consistently elicit this type of patient communication requires knowledge of effective communication strategies for a variety of patients, which can only be obtained through analysis of a large number of annotated interviews. Since manual examination and sequential pattern analysis of transcripts is a very time-consuming process, tailoring of MI interventions to particular populations can take years. Therefore, there is a need for informatics-based methods to facilitate the development of effective MI-based interventions, in general, and theoretically-grounded computational models to explore the mechanisms of MI's efficacy, in particular.

@@

\begin{figure}[!htb]
    \centering
    \includegraphics[width=0.50\textwidth]{figures/code_embed.eps}
    \caption{\textbf{T-SNE diagram of behavior codes drawn after 4000 steps with perplexity 30}.}
    \label{fig:code_embedding}
\end{figure}

@@

\subsection*{\textit{Most likely communication sequences}}
Table~\ref{tab:common_patterns} provides examples of typical patient-provider communication sequences that frequently appear in successful and unsuccessful motivational interviews. In successful sequences, we observed the provision of information using patient-centered communication (GINFO+) and structure session which is the counselor either explaining the therapeutic agenda or attempting to transition to a new topic or session content (SS). Sometimes, counselor acknowledged the clients' communication or an off topic comment (SO). We also observed that the affirm and OQECML+ are consistent with MI theory as an effect for eliciting positive change talk or commitment language. For unsuccessful sequences, It can be seen that providing advice using non-patient centered strategies (ADV-) leads to the ambivalence that is weighted against change (AMB-) which is heading in the wrong direction therapeutically. Questions phased to elicit negative change talk or commitment language leads to CHT- or CML-, which is consistent with the manual coding approach or analysis, or AMB-.

%It can be seen that most successful sequence starts with a summary of the discussion or structured statement. After that, if adolescents express positive change talk, the counselor immediately reflects on that to reinforce adolescent's intrinsic motivation about behavior change. On the other hand, asking open- or closed-ended questions with eliciting negative change talk can lead to negative change talk, even in the cases when adolescents were showing a positive tendency in their previous communication. This observation can be explained by adolescents quickly confused when reiterating negative information that undermines their motivation. Analyzing such cases will allow the counselors to determine the negative information that can be discussed during the interviews.
